---
title: "p8105_hw2_yh3676"
author: "Yuzhe Hu"
date: "2023-10-03"
output: github_document
---

```{r setup, message=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1
Clean `pols-month.csv` data
```{r step1, message=FALSE}
pols_month = 
  read_csv("./data/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year","month","day"), sep = "-") |>
  mutate(month = month.abb[as.numeric(month)],
         president =  ifelse(prez_gop == 1,"gop","dem")) |> 
  select(-prez_dem, -prez_gop, -day)
```

Clean `snp.csv` data
```{r step2, message=FALSE}
snp_df = 
  read_csv("./data/snp.csv") |>
  separate(date, into = c("month", "day", "year"), sep = "/") |>
  mutate(month = month.abb[as.numeric(month)]) |>
  mutate(year = ifelse(as.numeric(year) <= 16, paste0("20", year), paste0("19", year))) |>
  select(-day) |>
  select(year, month, everything())
```

Clean `unemployment.csv` data
```{r step3, message=FALSE}
unempl_df = 
  read_csv("./data/unemployment.csv") |>
  pivot_longer(Jan:Dec, names_to = "month", values_to = "unemployment rate") |> 
  rename(year = Year) |>
  mutate(year = as.character(year))
```

Merging `snp` into `pols`, and then merging `unemployment` into the result.
```{r}
# Merging `snp_df` into `pols_month`
merge_df_1 =
  left_join(pols_month, snp_df, by = c("year","month"))

# Then merging `unempl_df`:
merge_df = 
  left_join(merge_df_1, unempl_df, by = c("year","month"))
```

* The file`pol_month.csv` contains data on the number of democratic or republican national politicians at various points in time between 1947 and 2015. 
* The file `snp.csv` contains data related to Standard & Poor’s stock market index from 1950 to 2015. 
* The file `unemployment.csv` contains monthly percentage of unemployment data from 1948 to 2015.
* The resulting dataset is merged by the above three datasets, which comprises 822 observations and 11 variables. It covers a 68-year range from 1947 to 2015. Key variables in the resulting dataset are `year`,`month`,`president`,`close` and `unemployment rate`.


## Problem 2
Clean the _Mr. Trash Wheel_ sheet
```{r}
mr_wheel = 
  read_excel('./data/202309 Trash Wheel Collection Data.xlsx', sheet = 1, range = 'A2:N586') |>
  janitor::clean_names() |>
  mutate(year = as.numeric(year))
```

The data include a column for the (approximate) number of homes powered. This calculation is described in the `Homes powered note`, but not applied to every row in the dataset. Update the data to include a new `homes_powered` variable based on this calculation.

```{r homes_powered}
mr_wheel = mutate(mr_wheel, homes_powered = weight_tons * 500 / 30)
```

Use a similar process to import, clean, and organize the data for _Professor Trash Wheel_ and _Gwynnda_, and combine these with the _Mr. Trash Wheel_ dataset to produce a single tidy dataset. 

```{r, message=FALSE}
professor_wheel = 
  read_excel('./data/202309 Trash Wheel Collection Data.xlsx', sheet = 2, range = 'A2:M108') |>
  janitor::clean_names() |>
  mutate(homes_powered = weight_tons*500/30) |>
  mutate(type = 'professor trash wheel')

gwynnda_wheel = 
  read_excel('./data/202309 Trash Wheel Collection Data.xlsx', sheet = 4, range = 'A2:L157') |>
  janitor::clean_names() |>
  mutate(homes_powered = weight_tons*500/30) |>
  mutate(type = 'gwynnda trash wheel')

mr_wheel = mutate(mr_wheel, type = 'mr trash wheel')

## Combine the three datasets
Combine_wheel = 
  full_join(mr_wheel, professor_wheel) |> 
  full_join(gwynnda_wheel)
```

* The `Combine_wheel` is a collection of three datasets with `r nrow(Combine_wheel)` number of observations. `r nrow(mr_wheel)` observations from the _Mr. Trash Wheel_ dataset,  `r nrow(professor_wheel)` observations from the _Professor Trash Wheel_ dataset, and `r nrow(gwynnda_wheel)` observations from the _Gwynnda Trash Wheel_ dataset. For key variables, each observation contains the date (`month`, `year` and `date`), weight(`weight_tons`) and volume(`volume_cubic_yards`) of trash, number of different trash categories (e.g. `plastic_bags`, `cigarette_butts`), and which trash wheel it belongs to.
* The total weight of trash collected by Professor Trash Wheel is `r sum(pull(professor_wheel, weight_tons))` tons. 
* The total number of cigarette butts collected by Gwynnda in July of 2021 is `r gwynnda_wheel |> filter(year == 2021, month == 'July') |> pull(cigarette_butts) |> sum()`.

## Problem 3
Clean the dataset of baseline demographics.
```{r}
# import data and skip the first row
baseline_df = read.csv("./data/MCI_baseline.csv", skip = 1) |> 
  janitor::clean_names() |> 
# Change types of `sex` and `apoe4`
  mutate(sex = if_else(sex == 0, 'female', 'male'),
  apoe4 = if_else(apoe4 == 0, 'non-carrier', 'carrier')) |> 
# Remove unqualified observations
  filter(age_at_onset > current_age | age_at_onset == '.')
```
* Important steps in the import process including skipping the first row of the dataset, cleaning the variable names into tidy format, changing the form of showing variables `sex` and `apoe4`, and removing participants who do not meet the stated inclusion criteria.
* `r nrow(baseline_df)` participants were recruited (after removing unqualified participants), of these `r baseline_df |> filter(age_at_onset > current_age) |> nrow()` developed MCI.
* The average baseline age is `r mean(pull(baseline_df, current_age))`.
*  `r 100 * nrow(filter(baseline_df, sex == "female", apoe4 == "carrier")) / nrow(filter(baseline_df, sex == "female"))`% of women in the study are APOE4 carriers.


Clean _longitudinally observed biomarker values_ dataset
```{r amyloid_df, message=FALSE, warning=FALSE}
amyloid_df = 
  read_csv('./data/mci_amyloid.csv', skip = 1) |>
  janitor::clean_names() |>
  pivot_longer(
    baseline:time_8,
    names_to = 'time',
    values_to = 'amyloid_ratio'
  ) |>
  rename(id = study_id) |>
  mutate(amyloid_ratio = as.numeric(amyloid_ratio)) |> 
  mutate(time = case_match(time,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8
    ))
```
* The dataset `amyloid_df` is imported by skipping the first row. Use `pivot_longer` to transform the dataset from wide to long format. The column name `study_id` is changed to `id` to match with the `baseline_df` dataset. The data type of `amyloid_ratio` is converted to numeric through `mutate`. 
* The dateset `amyloid_df` has `r ncol(amyloid_df)` variables, with variale names: `r names(amyloid_df)`.

Check whether some participants appear in only the baseline or amyloid datasets. 
```{r}
##Select the id
id_baseline = unique(pull(baseline_df, id))
id_amyloid = unique(pull(amyloid_df, id))
#Id of participants that only appear in `MCI_baseline.csv`
id_baseline_only = id_baseline[!(id_baseline %in% id_amyloid)]
#Id of participants that only appear in `mci_amyloid.csv`
id_amyloid_only = id_amyloid[!(id_amyloid %in% id_baseline)] 
```

* A total of `r length(setdiff(pull(baseline_df, id), pull(amyloid_df, id)))` participants appear in only the baseline dataset, and their id are `r id_baseline_only`. A total of `r length(setdiff(pull(amyloid_df, id), pull(baseline_df, id)))` participants appear in only the amyloid dataset, and their id are `r id_amyloid_only`. 

Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and export the result as a CSV.
```{r}
#Combine the data set
combined_df = inner_join(baseline_df, amyloid_df, by = "id")
#Export the combined data
write_csv(combined_df,"./data/combined_df.csv")
```
* The resulting dataset `combined_df` contains the data of `r combined_df |> pull(id) |> unique() |> length()` participants who appear in both datasets, including their basic demographic information (`id`,  `current_age`, `sex`, `education` ), APOE4 carrier status (`apoe4`), `age_at_onset` of MCI and their amyloid β 42/40 ratio (`amyloid_ratio`) at measurement time (`time`). `r combined_df |> filter(apoe4 == "carrier") |> pull(id) |> unique() |> length()` of the participants are APOE4 carriers, and `r combined_df |> filter(apoe4 == "carrier", age_at_onset > 0) |> pull(id) |> unique() |> length()` of them developed MCI. `r combined_df |> filter(apoe4 == "non-carrier") |> pull(id) |> unique() |> length()` of the participants are non-carriers, and `r combined_df |> filter(apoe4 == "non-carrier", age_at_onset > 0) |> pull(id) |> unique() |> length()` of them developed MCI.





