---
title: "p8105_hw2_yh3676"
author: "Yuzhe Hu"
date: "2023-10-03"
output: github_document
---

```{r setup, message=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1
Clean `pols-month.csv` data
```{r step1, message=FALSE}
pols_month = 
  read_csv("./data/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year","month","day"), sep = "-") |>
  mutate(month = month.abb[as.numeric(month)],
         president =  ifelse(prez_gop == 1,"gop","dem")) |> 
  select(-prez_dem, -prez_gop, -day)
```

Clean `snp.csv` data
```{r step2, message=FALSE}
snp_df = 
  read_csv("./data/snp.csv") |>
  separate(date, into = c("month", "day", "year"), sep = "/") |>
  mutate(month = month.abb[as.numeric(month)]) |>
  mutate(year = ifelse(as.numeric(year) <= 16, paste0("20", year), paste0("19", year))) |>
  select(-day) |>
  select(year, month, everything())
```

Clean `unemployment.csv` data
```{r step3, message=FALSE}
unempl_df = 
  read_csv("./data/unemployment.csv") |>
  pivot_longer(Jan:Dec, names_to = "month", values_to = "unemployment rate") |> 
  rename(year = Year) |>
  mutate(year = as.character(year))
```

Merging `snp` into `pols`, and then merging `unemployment` into the result.
```{r}
# Merging `snp_df` into `pols_month`
merge_df_1 =
  left_join(pols_month, snp_df, by = c("year","month"))

# Then merging `unempl_df`:
merge_df = 
  left_join(merge_df_1, unempl_df, by = c("year","month"))
```

* The file`pol_month.csv` contains data on the number of democratic or republican national politicians at various points in time between 1947 and 2015. 
* The file `snp.csv` contains data related to Standard & Poor’s stock market index from 1950 to 2015. 
* The file `unemployment.csv` contains monthly percentage of unemployment data from 1948 to 2015. * The resulting dataset is merged by the above three datasets, which comprises 822 observations and 11 variables. It covers a 68-year range from 1947 to 2015. Key variables in the resulting dataset are `year`,`month`,`president`,`close` and `unemployment rate`.


## Problem 2
Clean the _Mr. Trash Wheel_ sheet
```{r}
mr_wheel = 
  read_excel('./data/202309 Trash Wheel Collection Data.xlsx', sheet = 1, range = 'A2:N586') |>
  janitor::clean_names() |>
  mutate(year = as.numeric(year))
```

The data include a column for the (approximate) number of homes powered. This calculation is described in the `Homes powered note`, but not applied to every row in the dataset. Update the data to include a new `homes_powered` variable based on this calculation.

```{r homes_powered}
mr_wheel = mutate(mr_wheel, homes_powered = weight_tons * 500 / 30)
```

Use a similar process to import, clean, and organize the data for _Professor Trash Wheel_ and _Gwynnda_, and combine these with the _Mr. Trash Wheel_ dataset to produce a single tidy dataset. 

```{r, message=FALSE}
professor_wheel = 
  read_excel('./data/202309 Trash Wheel Collection Data.xlsx', sheet = 2, range = 'A2:M108') |>
  janitor::clean_names() |>
  mutate(homes_powered = weight_tons*500/30) |>
  mutate(type = 'professor trash wheel')

gwynnda_wheel = 
  read_excel('./data/202309 Trash Wheel Collection Data.xlsx', sheet = 4, range = 'A2:L157') |>
  janitor::clean_names() |>
  mutate(homes_powered = weight_tons*500/30) |>
  mutate(type = 'gwynnda trash wheel')

mr_wheel = mutate(mr_wheel, type = 'mr trash wheel')

## Combine the three datasets
Combine_wheel = 
  full_join(mr_wheel, professor_wheel) |> 
  full_join(gwynnda_wheel)
```

* The `Combine_wheel` is a collection of three datasets with `r nrow(Combine_wheel)` number of observations. `r nrow(mr_wheel)` observations from the _Mr. Trash Wheel_ dataset,  `r nrow(professor_wheel)` observations from the _Professor Trash Wheel_ dataset, and `r nrow(gwynnda_wheel)` observations from the _Gwynnda Trash Wheel_ dataset. For key variables, each observation contains the date (`month`, `year` and `date`), weight(`weight_tons`) and volume(`volume_cubic_yards`) of trash, number of different trash categories (e.g. `plastic_bags`, `cigarette_butts`), and which trash wheel it belongs to.
* The total weight of trash collected by Professor Trash Wheel is `r sum(pull(professor_wheel, weight_tons))` tons. 
* The total number of cigarette butts collected by Gwynnda in July of 2021 is `r gwynnda_wheel |> filter(year == 2021, month == 'July') |> pull(cigarette_butts) |> sum()`.



