---
title: "p8105_hw2_yh3676"
author: "Yuzhe Hu"
date: "2023-10-03"
output: github_document
---

```{r setup, message=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1
Clean `pols-month.csv` data
```{r step1, message=FALSE}
pols_month = 
  read_csv("./data/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, into = c("year","month","day"), sep = "-") |>
  mutate(month = month.abb[as.numeric(month)],
         president =  ifelse(prez_gop == 1,"gop","dem")) |> 
  select(-prez_dem, -prez_gop, -day)
```

Clean `snp.csv` data
```{r step2, message=FALSE}
snp_df = 
  read_csv("./data/snp.csv") |>
  separate(date, into = c("month", "day", "year"), sep = "/") |>
  mutate(month = month.abb[as.numeric(month)]) |>
  mutate(year = ifelse(as.numeric(year) <= 16, paste0("20", year), paste0("19", year))) |>
  select(-day) |>
  select(year, month, everything())
```

Clean `unemployment.csv` data
```{r step3, message=FALSE}
unempl_df = 
  read_csv("./data/unemployment.csv") |>
  pivot_longer(Jan:Dec, names_to = "month", values_to = "unemployment rate") |> 
  rename(year = Year) |>
  mutate(year = as.character(year))
```

Merging `snp` into `pols`, and then merging `unemployment` into the result.
```{r}
# Merging `snp_df` into `pols_month`
merge_df_1 =
  left_join(pols_month, snp_df, by = c("year","month"))

# Then merging `unempl_df`:
merge_df = 
  left_join(merge_df_1, unempl_df, by = c("year","month"))
```

* The file`pol_month.csv` contains data on the number of democratic or republican national politicians at various points in time between 1947 and 2015. The file `snp.csv` contains data related to Standard & Poor’s stock market index from 1950 to 2015. The file `unemployment.csv` contains monthly percentage of unemployment data from 1948 to 2015. The resulting dataset is merged by the above three datasets, which comprises 822 observations and 11 variables. It covers a 68-year range from 1947 to 2015. Key variables in the resulting dataset are `year`,`month`,`president`,`close` and `unemployment rate`.
